### 2019-09-17 reports/thesis/chapters/3_compoundPoisson.tex
Adjust Fisher information matrix in compound Poisson

The EM algorithm \citep{dempster1977maximum} treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{Y_1,Y_2,Y_3,\dotsc, Y_n\right\}$ be realisations of $Y$. Define the joint log-likelihood to be
\begin{multline*}
  \ln L(\lambda,\alpha,\beta;X,Y)
  =
  \sum_{i=1}^n
  \left[
    \mathbb{I}(x_i=0)
    \ln
    \prob(Y=0)
  \right.
  \\
  \left.
    +
    \mathbb{I}(x_i>0)
    \ln
    \left[
      p_{X|Y}(x_i|Y_i)\prob(Y=Y_i)
    \right]
  \right]
\end{multline*}
so that
\begin{multline}
  \ln L(\lambda,\alpha,\beta;X,Y)=
  \sum_{i=1}^n
  \left[
    -\mathbb{I}(x_i=0)
    \lambda
  \right.
  \\
  \left.+
    \mathbb{I}(x_i>0)
    \left(
      Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
    \right.
  \right.
  \\
  \left.
    \left.  
      - \lambda + Y_i \ln \lambda - \ln(Y_i!)
    \right)
  \right]
  \ .
\end{multline}
The estimators were found by optimising the joint log-likelihood. This was done iteratively by estimating the $Y$s given the $X$s and parameters (E step), followed by estimating the parameters given the $Y$s and $X$s (M step) until some convergence conditions were met.

@@

The variance of the M step estimators can be obtained by using the Cram\'er-Rao lower bound \citep{rao1945information, cramer1946mathematical} from the Fisher's information matrix \citep{fisher1922on} defined as
\begin{equation}
  \matr{I} = -\expectation\left[
    \nabla_{\lambda,\alpha,\beta}\nabla_{\lambda,\alpha,\beta}\T \ln L(\lambda,\alpha,\beta;X,Y)
  \right]
\end{equation}
where
\begin{equation}
  \nabla_{\lambda, \alpha,\beta}=
  \begin{pmatrix}
    {\partial}/{\partial \lambda}
    \\
    {\partial}/{\partial \alpha}
    \\
    {\partial}/{\partial \beta}
  \end{pmatrix}
  \ .
\end{equation}
The calculation is similar to the derivatives of $T$ using the same approximations. Using the fact that $\expectation[Y]=\lambda$ and another approximation $\expectation\left[\psi^{(r)}(\alpha Y)\right]\approx\psi^{(r)}(\alpha \lambda)$, then
\begin{equation}
  \matr{I}=
  \begin{pmatrix}
    \dfrac{n}{\lambda} & 0 & 0 \\
    0 & (\lambda^2+\lambda)\psi'(\alpha\lambda)+2\alpha\lambda^2\psi''(\alpha\lambda)+\frac{1}{2}\alpha^2\lambda^3\psi'''(\alpha\lambda) & -\dfrac{n\lambda}{\beta}\\
    0 & -\dfrac{n\lambda}{\beta} & \dfrac{n\alpha\lambda}{\beta^2}
  \end{pmatrix}
  \ .
\end{equation}
The Cram\'er-Rao lower bound is then
\begin{equation}
  \cov\left[
    \begin{pmatrix}
      \widehat{\lambda}\\\widehat{\alpha}\\\widehat{\beta}
    \end{pmatrix}
  \right]
  =
  \matr{I}^{-1}
  \ .
\end{equation}

>>>>

The EM algorithm \citep{dempster1977maximum} treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{Y_1,Y_2,Y_3,\dotsc, Y_n\right\}$ be realisations of $Y$. Define the joint log-likelihood to be
\begin{multline*}
  \ln L(\lambda,\alpha,\beta;X,Y)
  =
  \sum_{i=1}^n
  \left[
    \mathbb{I}(x_i=0)
    \ln
    \prob(Y=0)
  \right.
  \\
  \left.
    +
    \mathbb{I}(x_i>0)
    \ln
    \left[
      p_{X|Y}(x_i|Y_i)\prob(Y=Y_i)
    \right]
  \right]
\end{multline*}
so that
\begin{multline}
  \ln L(\lambda,\alpha,\beta;X,Y)=
  \sum_{i=1}^n
  \left[
    -\mathbb{I}(x_i=0)
    \lambda
  \right.
  \\
  \left.+
    \mathbb{I}(x_i>0)
    \left(
      Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
    \right.
  \right.
  \\
  \left.
    \left.  
      - \lambda + Y_i \ln \lambda - \ln(Y_i!)
    \right)
  \right]
  \ .
  \label{eq:compoundPoisson_jointLogLikelihood}
\end{multline}
The estimators were found by optimising the joint log-likelihood. This was done iteratively by estimating the $Y$s given the $X$s and parameters (E step), followed by estimating the parameters given the $Y$s and $X$s (M step) until some convergence conditions were met.

@@

The variance of the M step estimators can be obtained by using the Cram\'er-Rao lower bound \citep{rao1945information, cramer1946mathematical} from the Fisher's information matrix \citep{fisher1922on} defined as
\begin{equation}
  \matr{I} = -\expectation\left[
    \nabla_{\lambda,\alpha,\beta}\nabla_{\lambda,\alpha,\beta}\T \ln L(\lambda,\alpha,\beta;X,Y)
  \right]
\end{equation}
where
\begin{equation}
  \nabla_{\lambda, \alpha,\beta}=
  \begin{pmatrix}
    {\partial}/{\partial \lambda}
    \\
    {\partial}/{\partial \alpha}
    \\
    {\partial}/{\partial \beta}
  \end{pmatrix}
  \ .
\end{equation}

Most of the calculations for the elements of the matrix are similar to the derivatives of $T$. For the $\alpha$ component, taking derivatives of the joint log-likelihood (Equation \eqref{eq:compoundPoisson_jointLogLikelihood}) obtains
\begin{equation}
\dfrac{\partial\ln L}{\partial\alpha} = \sum_{i=1}^n\mathbb{I}(x_i>0)
\left[
  Y_i\ln\beta - \psi(Y_i\alpha)Y_i+Y_i\ln x_i
\right]
\end{equation}
\begin{equation}
\dfrac{\partial^2\ln L}{\partial\alpha^2} = \sum_{i=1}^n\mathbb{I}(x_i>0)
\left[
  -\psi'(Y_i\alpha)Y_i^2
\right] \ .
\end{equation}
The expectation of the second-order derivative is
\begin{equation}
-\expectation\left[\dfrac{\partial^2\ln L}{\partial\alpha^2}\right]
=
n\left(1-\euler^{-\lambda}\right)
\expectation\left[
  \psi'(Y\alpha)Y^2
\right]
\end{equation}
where the factor $\left(1-\euler^{-\lambda}\right)$ is the proportion of data where $x_i>0$. $\expectation\left[\psi'(Y\alpha)Y^2\right]$ can be approximated. Let $g(Y)=\psi'(Y\alpha)Y^2$, then
\begin{equation}
\expectation\left[\psi'(Y\alpha)Y^2\right]
\approx
\psi'(\lambda\alpha)\lambda^2
+\dfrac{\lambda}{2}g''(\lambda) \ .
\end{equation}
The derivatives of $g(Y)$ are
\begin{align}
g'(Y)&=\psi''(Y\alpha)Y^2\alpha + 2\psi'(Y\alpha)Y
\\
g''(Y)&=Y^2\alpha^2\psi'''(Y\alpha)+4Y\alpha\psi''(Y\alpha)+2\psi'(Y\alpha)\ .
\end{align}
Therefore,
\begin{multline}
-\expectation\left[\dfrac{\partial^2\ln L}{\partial\alpha^2}\right]=n\left(1-\euler^{-\lambda}\right)
\\
\times\left[(\lambda^2+\lambda)\psi'(\lambda\alpha)+2\lambda^2\alpha\psi''(\lambda\alpha)+\dfrac{\lambda^3\alpha^2}{2}\psi'''(\lambda\alpha)\right]
\end{multline}

The Fisher's information matrix is
\begin{multline}
  \matr{I}=
  n
  \left(\begin{matrix}
    \dfrac{1}{\lambda} & 0\\
    0 & \left(1-\euler^{-\lambda}\right)\left[(\lambda^2+\lambda)\psi'(\lambda\alpha)+2\lambda^2\alpha\psi''(\lambda\alpha)+\dfrac{\lambda^3\alpha^2}{2}\psi'''(\lambda\alpha)\right]\\
    0 & -\dfrac{\left(1-\euler^{-\lambda}\right)\lambda}{\beta}
  \end{matrix}\right.
  \\
  \left.\begin{matrix}
  \vphantom{\dfrac{1}{\lambda}}0\\
  -\dfrac{\left(1-\euler^{-\lambda}\right)\lambda}{\beta}\\
  \dfrac{\left(1-\euler^{-\lambda}\right)\alpha\lambda}{\beta^2}
  \end{matrix}\right)
  \ .
\end{multline}
The Cram\'er-Rao lower bound can be obtained using
\begin{equation}
  \cov\left[
    \begin{pmatrix}
      \widehat{\lambda}\\\widehat{\alpha}\\\widehat{\beta}
    \end{pmatrix}
  \right]
  =
  \matr{I}^{-1}
  \ .
\end{equation}
